### 2010-11-29 trunk/htdocs/porous.tex
Seaparte file for porous media stuff
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1171 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.14 2010-04-30 11:00:50 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Note to Readers}

This document is intended as a starting point for a full-blown
specification of the next version of the Ludwig code, and everything
to do with it. It is (manifestly) not complete.

\subsection{Some History}

Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
Ludwig.

\subsection{Aims}

We aim to provide a robust and portable code, written in C, which
can be used to perform serial and scalable parallel simulations of
complex fluid systems based around the Lattice Boltzmann method.

\subsection{Availability}

We now have a NeSCForge project account available at

\texttt{http://forge.nesc.ac.uk/projects/ludwig/}

This provides controlled access to known
persons and provide CVS, bug tracking, and so on. It is
\textbf{essential} that people sign up to this, as we must
move to a single supported version.

The NeSCForge CVS is not available to the public.

\subsection{Download}

\section{User Input}

By default the executable will look for an input file at run time.
By default, this file is named \texttt{input} and should be in the
current working directory. The alternative is to specify the name
of the file explicitly on the command line:
\begin{verbatim}
$ ./Ludwig.exe input_file
\end{verbatim}

The contents of the input file are made up of \textit{key value}
pairs with control the run-time behaviour. Blank lines and lines
begining with \# are ignored as comments. The following describes
the effect of various keys. Most keys have a default value which
will be used by the code if the corresponding key is not present
(or commented out) in the input file.

\subsection{Basic run parameters}

\inputkey{N\_cycles}

The number of lattice Boltzmann time steps to execute. Default
value:

\texttt{N\_cycles 0}

\inputkey{t\_start}

By default the code will start from time step $t = 0$. If you 
wish to restart from a previously saved configuration, set
the appropriate value for \texttt{t\_start}. The code will
execute \texttt{N\_cycles} steps starting from this point.

\inputkey{input\_config}

The file name stub for input configuration required for restarts.
For example, to restart from the file config.out100 previously
produced at the end of a 100 step run use
\begin{verbatim}
t_start 100
input_config config.out100
\end{verbatim}

If you are using parallel I/O, the filename stub is required; the
file extensions are computed automatically.

\inputkey{size}

Controls the number of lattice points in the $(x, y, z)$ directions,
respectively. Default value:

\texttt{size 64\_64\_64}

\inputkey{grid}

In parallel, this sets the processor Cartesian decomposition, i.e.,
the number of processors in each dimension of the Cartesian communicator.
The default value may be implementation dependent as it is that returned
by \texttt{MPI\_Dims\_create()}. For
example:
\begin{verbatim}
size 64_64_64
grid 4_2_1
\end{verbatim}
gives a total (physical) system size of 64 lattice sites in each
vdirection, decomposed across 4 processors in the x-direction, 2 in
the y-direction and 1 (no decomposition) in the z-direction. The
total number of processors must therefore be 8. The local domain
size per processor is then 16x32x64.

\subsection{Fluid properties}

\inputkey{viscosity}

Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
(i.e., realaxation time equal to unity).

\inputkey{viscosity\_bulk}

Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
This can be use if you are concerned with incompressiblity
violations, in which case $\zeta >> \eta$ can be useful. The
default value is $\zeta = \eta$.

\inputkey{temperature}

The LB 'temperature' using fluctuating hydrodynamics. Safe values
are $0.0001 > kT > 0$.

\inputkey{rho0}

Sets the mean density of the fluid. The default value is unity, and
it is recommended that the value should not be changed.

\subsection{Binary fluid parameters}

The are currently three choices of free energy relating to
various systems.

\subsubsection{Symmetric binary fluid}

\inputkey{free\_energy symmetric}

The following applies for the binary fluid problem with
compositional order parameter $\phi$ and free energy
(excluding the term in the density $\rho$):
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
\end{equation}

This is described in some detail
by Kendon et al \cite{viv}. The first two terms represent the bulk
contribution, whereas the term in $\kappa$ penalisies curvature in
the interface.

\inputkey{A}

The parameter $A$. Note $A < 0$.

\inputkey{B}

The parameter $B$. Note $B = -A$ for common usage, although this is
not enforced.

\inputkey{K}

The parameter $\kappa$, which is positive.

\subsubsection{Brazovskii free energy}

This is similar to the symetric, but with one extra term which
requires description.
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
\end{equation}

\subsubsection{Surfactant model free energy}

\inputkey{free\_energy surfactant}

This is for use with a two order parameter model the first of
which is the composition, as for the symmetric free energy,
while the second $\psi$ represents surfactant concentration.
The free energy density is made up of a number of terms:

\begin{equation}
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
\end{equation}
is the standard symmetric term related to the composition as above;

\begin{equation}
D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
\end{equation}
represents the energy of surfactant in the bluk phase with surfactant
concentration $0 < \psi < 1$ with a single parameter $D$;
\begin{equation}
-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
\end{equation}
where the term in $\epsilon$ represents the energy reduction by
adsorbing surfactant at the interface and the term in $\beta$
provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
\begin{equation}
+{\textstyle \frac{1}{2}} W \psi \phi^2
\end{equation}
which is included to stabilise the interface \cite{theissengompper}.

The corresponding keys for the input file are:

\inputkey{D}

is the bulk surfactant contribution parameter

\inputkey{epsilon}

is the parameter $\epsilon$

\inputkey{W}

is the parameter $W$

\inputkey{beta}

is the value of the parameter $\beta$ (not implemented yet).

\inputkey{phi\_b}

the initial, uniform, background surfactant concentration $\psi_b$.

\subsection{Other Binary Fluid Parameters}

\inputkey{mobility}

Sets the compositional order parameter mobility $M$ (uniform in space)
which appears in the Cahn-Hilliard equation.

\inputkey{mobilty\_psi}

Sets the (uniform) mobility for the surfactant as appropriate for
the Cahn-Hilliard equation.

\inputkey{phi0}

The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{phi\_initialisation}

Determines how the compositional order parameter is initialisated
at the start of the run. If set to \texttt{spinodal} the value
is set to $\phi_0$ as set above plus or minus a random noise, the
magnitude of which is set by the value of the \texttt{noise} key.
If set to \texttt{block} a one-dimensional profile is set up in
the $z$-direction representing two blocks of fluid with
$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
appropriate width. The $\phi = -1$ section is in the middle.

\inputkey{noise}

The magnitude of the initial fluctuations in $\phi$ used to
initiate spinodal decompsotion.

\subsection{Colloidal particles}

There are quite a few options here.

\inputkey{colloid\_init}

This determines how the code initialises particles (or not). The
default is to have no particles (value \texttt{no\_colloids}). There
are two other options at the moment.

\texttt{colloid\_init    fixed\_number\_monodisperse}

initialises a fixed number of particles with randomly selected
positions. Note that there is no gaurentee that the particles
do not overlap. The number of particles is determined by
the value of the key \texttt{colloid\_no}.

\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}

initialises a given volume fraction of particles on a (roughly)
bcc lattice. The particles will not overlap. The volume fraction
is set using the key \texttt{colloid\_vf}.

\inputkey{colloid\_a0}

Sets the nominal radius of the particle on the lattice $a_0$. This is
used to construct the links for BBL. The smallest acceptable LB
particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.

\inputkey{colloid\_ah}

Sets the hydrodynamic radius of the particles $a_h$. For all
reasonable applications choose $a_h = a_0$.

\inputkey{colloid\_mc\_steps}

It is possible to have the code follow the initialisation of
particle positions by a number of Monte Carlo steps. This can
help to remove overlaps (fixed\_number\_monodisperse) or
melt the lattice (fixed\_volume\_fraction\_monodisperse).

\inputkey{colloid\_gravity}

Sets a body force on each paarticle. Useful, for example, for
sedimentation. For example,

\inputkey{cell\_list\_min}

This sets the minimum allowable width of the cell list used to
store the particles. This should be as small as possible for
efficiency purposes but must be large enough to capture all
possible interactions.

\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}

gives a constant force in the negative $z$-direction on each particle.
The code automatically computes the compensating body force on fluid nodes
required to give no net change in the total momentum of the system at
each time step.

\subsection{Active particles}

Not in general use yet.

\subsection{Particle-particle interactions}

\subsection{Boundary conditions}

Solid planar boundary walls can be added at the extremities of the
system (i.e., the volume of fluid is extactly that specified by
the \texttt{size} keyword). The wall can be in one, two or all three
coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
or a 'box' geometry, respectively.

The boundary walls are switched on via the keyword value pair

\inputkey{boundary\_walls\_on [no | yes]}

If \texttt{yes} then the geometry is specified with the keyword
as one of

\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}

Note that these are the only valid selections. \texttt{1\_1\_0}
gives periodic boundaries in the $x-$ and $y-$directions, with
walls at either edge in the $z-$direction. \texttt{1\_0\_0}
gives periodic conditions in the $x-$ direction only, while
\texttt{0\_0\_0} is a fully enclosed box.

In the case of the slab geometry, two further values are used so
that shear may be imparted to the system:

\inputkey{boundary\_speed\_bottom}

\inputkey{boundary\_speed\_top}

These values set the $x-$component of the wall velocity at
$z = z_{min}$ and $z = z_{max}$, respectively. Like all
velocities, these values must respect the low Mach number
constraint $u_x < c_s$.

\subsection{Input/Output and Parallel I/O}

\inputkey{freq\_statistics}

sets the frequency for reporting general information from
the code. (In general, the function
 \texttt{is\_statistics\_step()} returns true at this frequency.

\inputkey{freq\_measure}

sets the frequency of recording of colloid/order parameter output.
(In general, the function \texttt{is\_measurement\_step()} returns
true at this frequency.)

\inputkey{freq\_config}

set frequency of full configuration (checkpoint) dumps suitable
for restart.

\inputkey{freq\_phi}

set frequency of order parameter output.

\inputkey{phi\_format}

either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.

\inputkey{freq\_vel}

set frequancy of velocity output.

\inputkey{vel\_format}

either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.

\inputkey{config\_at\_end}

enforces full configuration dump at the end of execution. The
default is 'yes'.

\subsection{Lees-Edwards planes}

\inputkey{N\_LE\_plane}

sets the total number of Lees-Edwards planes. Default is zero.
The placing is as followings. The number of planes $n$ must
divide the lattice size in the $x$-direction to give an integer
$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.

\inputkey{LE\_plane\_vel}

sets the velocity of each plane relative to the lattice. All planes
have the same, constant, velocity.

\inputkey{LE\_init\_profile}

if set to 1, the fluid velocity is initialise to reflect a steady
state shear flow appropriate for the number of planes at the
given velocity. If set to zero, the fluid is initialised with
zero velocity.

@@

\subsection{Miscellaneous}

\inputkey{random\_seed}

sets the random number generator seed.

\subsection{Dealing with the Parallel I/O files}

When running in parallel, output of lattice-based (and particle)
quantities takes place in parallel. As a consequence, the order
of the output in the files is not in the same order as would be
the case in serial. This means the parallel output must be
manipulated into the correct order before any analysis.

Parallel I/O subdivides the processors into one or more groups,
each of which write data to a separate file. Infomation on the
content of each of these files is contained in `meta-data' file
for each quantity, which are produced automatically at the start
of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
If there are 8 I/O groups, the files will be named \texttt{008-001}
to \texttt{008-008} for the different groups.

The actual data is stored for each relevant time step as, e.g.,
\texttt{phi-000100.001-001} where the extention refers to the
I/O. To recombined a single, or multiple, data files into a
single file in the correct order, a utility program
\texttt{extract.c} is provided.

The exectutable takes two arguments, e.g.,

\inputkey{./extract phi.001-001.meta phi-000100.001-001}

where  the first is the first meta-data file in the relevant series,
and the second is the first daa file for the relevant time step. The
program will combine the relevant quantities in the correct order
and produce a singel new file \texttt{phi-000100} (in the above case).

\subsection{Restarting}

When restarting in parallel, the processor decomposition must be
preserved. An attempt to change the processor decomposition will
cause the configuration files to be read in the incorrect order,
and results will be wrong.

\section{The Lattice Boltzmann Method}

We review here the lattice Boltzmann method applied to single
phase problems.

\subsection{The Navier Stokes Equation}

We seek to solve the isothermal Navier-Stokes equations which, often
written in vector form, express mass conservation
\begin{equation}
\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
\label{eq_mass1}
\end{equation}
and the conservation of momentum
\begin{equation}
\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
-\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
\label{eq_momentum1}
\end{equation}
Equation~\ref{eq_mass1} expresses the local rate of change of the
density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
Equation~\ref{eq_momentum1} expresses Newton's second law for
momentum, where the terms on the right hand side represent the
force on the fluid.

For this work, it is more convenient to rewrite these equations
in tensor notation, where Cartesian coordinates ${x,y,z}$ are
represented by indices $\alpha$ and $\beta$, viz
\begin{equation}
\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
\end{equation}
and
\begin{equation}
\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
= -\nabla_\alpha p
+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
\end{equation}
Here, repeated indices are understood to be summed over.
The conservation law is seem better if the forcing terms of the
right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
so that
\begin{equation}
\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
\end{equation}
In this case the expanded expression for the stress tensor is
\begin{equation}
\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
\end{equation}
where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
equations have 10 degrees of freedom (or hydrodynamic modes) being
$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.

\subsection{The Distribution Function and its Moments}

In lattice Boltzmann, the density and velocity of the continuum fluid
are replaced by their discrete counterparrts, the distribution function
$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
It is possible to relate the hydrodynamic modes to distribution
function via its moments, so
\begin{equation}
\rho = \sum_i f_i,  \,
\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
\end{equation}
here, the index of the summation is over the number of discrete
velocities used $n$, which in three dimensions is generally
denoted by D3Q$n$. We restrict our discussion to D3Q15 and
D3Q19 (see below).

In general, the number of modes supported in $n$ and these can
be written as
\begin{equation}
M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
\end{equation}
where $m_i^a$ is the corresponding eigenvector of the collision
maxtrix (see section blah). For example, if $m_i^a = 1$ then the
mode $M^a$ is the density. Correspondingly, the distribution
function can be related to the modes $M^a$ via
\begin{equation}
f_i() = \sum_a w_i m_i^a M^a() N^a
\end{equation}

\subsubsection{Hydrodynamic modes}

\subsubsection{D3Q15 ghost modes}

\subsubsection{D3Q19 ghost modes}

\subsection{The Lattice Boltzmann Equation}

\subsection{The Collision}

\subsection{External Body Forces}

\subsection{Fluctuating LBE}

It is possible \cite{adhikari2005} to simulate fluctuating
hydrodynamics for an isothermal fluid via the inclusion of
a fluctuating stress $\sigma_{\alpha\beta}$:
\begin{equation}
\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
\end{equation}
The fluctuation-dissipation theorem relates the magnitude of this
random stress to the isothermal temperature and the viscosity.

In the LBE, this translates to the addition of a random contribution
$\xi_i$ to the distribution at the collision stage, so that
\begin{equation}
\ldots + \xi_i.
\end{equation}

For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
i.e., those with dissipation, the fluctuating part may be written
\begin{equation}
\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
\end{equation}
where $\hat{\xi}^a$ is a noise termwhich has a variance determined
by the relaxation time for given mode
\begin{equation}
\left< \hat{\xi}^a \hat{\xi}^b \right> =
\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
\left< \delta M^a \delta M^b \right>.
\label{eq_fvar}
\end{equation}

\subsubsection{Fluctuating stress}

For the stress, the random contribution to the distributions is
\begin{equation}
\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
\end{equation}
where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
variates drawn from a Gaussian distribution with variance given
by equation~\ref{eq_fvar}. In the case that the shear and bulk
viscosities are the same, i.e., there is a single relaxation
time, then the variances of the six independent components of
the matrix are given by
\begin{equation}
\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
\frac{2\tau + 1}{\tau^2}
(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
\end{equation}

\subsubsection{Fluctuating ghost modes: D3Q15}

\subsubsection{Fluctuating ghost modes: D3Q19}

\subsection{The Propagation}

\subsection{Solid bodies and Bounce-Back on Links}

A very general method for the representation of solid objects
within the LB approach was put forward by Ladd \cite{l94a, l94b}.
Solid objects (of any shape) are defined by a boundary surface
which intersects some of the velocity vectors $\mathbf{c}_i$
joining lattice nodes. Sites inside are designated solid, while
sites outside remain fluid. The correct boundary condition is
defined by identifying \textit{links} between fluid and solid
sites, which allows those elements of the distribution which would
cross the boundary at the propagation step to be ``bounced-back''
into the fluid. This bounce-back on links is an efficient method
to obtain the  correct hydrodynamic interaction between solid
and fluid.

\subsubsection{Boundary walls}

\subsection{Colloids}

\subsubsection{General}

Colloidal particles are assumed to be spherical with a geometrical
centre $\mathbf{r}_c$, which is also the centre of mass. The
centre is allowed to move continuously across the lattice
with velocity $\mathbf{U}$; the particle has an angular velocity
$\mathbf{\Omega}$. The surface of the colloid is defined by an
input radius, $a_0$, which determines which lattice nodes are
inside or outside the colloid. (The hydrodynamic properties of
the colloid are specified by a different radius $a_h$ --- more
of this later.) The boundary links are then the set of vectors
joining lattice nodes which intersect the spherical surface
$\{\mathbf{c}_b\}$. A schematic picture is presented in
Figure~\ref{fig_coll2}. Note that a lattice node exactly at
the solid-fluid interface is defined to be outside the colloid.

In the original approach of Ladd, fluid occupied nodes both inside
and outside the particle. The effect of the ``internal fluid'' is
known to be restricted to short time scales (compared to the
characteristic time $a_0^2/\nu$), on which the fluid inside the
particle relaxes to a solid body rotation \cite{heemels}. There
are a number of problems related to the internal fluid, so we
use fully solid particles.

Nguyen and Ladd.

For each boundary link, it is useful to define a boundary vector,
$\mathbf{r}_b$, which connects the centre of the colloid to the
appropriate boundary node (note that the length of this vector is
not necessarily equal to the input radius of the colloid).

\begin{figure}[tb]
\begin{center}
\includegraphics{xfig/colloidhalflinksnew.eps}
\end{center}
\caption{Boundary links for a colloidal particle in two dimensions
with D2Q9 (half the particle is shown). Links join fluid sites (red)
to solid sites (green) and
intersect the circular shell radius $a_0$. Boundary nodes (circles)
lie exactly half way between pairs of lattice nodes. The generalisation
to three dimensions is straightforward.}
\label{fig_coll2}
\end{figure}

A \textit{boundary node} halfway between fluid nodes which are joined
by a boundary link. The position of the boundary node is always
exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
\Delta t$) regardless of the actual position of the intersection
of the colloid surface and the boundary link.
The colloid therefore has a discrete representation which becomes a
better approximation to the sphere as the input radius becomes larger;
this approximation is known to be reasonably good for $a_0 > 5\Delta x$.

\subsubsection{Bounce-back on links}

The standard boundary condition required for the solid-fluid
interface of a moving particle is described by Ladd \cite{l94b},
and is generally refered to as bounce-back on links (BBL).
A boundary link is defined as joining a node $\mathbf{r}$
inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
If the post-collision distributions are denoted by $f^\ast$, then
the distributions must be reflected at the solid surface so that
\begin{equation}
\label{eq:bbl1}
f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
\end{equation}
where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
at the boundary
\begin{equation}
\label{eq:ub}
\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
\end{equation}
is determined by the particle linear velocity $\mathbf{U}$ and angular
velocity $\mathbf{\Omega}$. The change in momentum described by

The force exerted on a
single link is
\begin{equation}
\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
\label{eq:fb}
\end{equation}
with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.

The total hydrodynamic force on the particle is then found by taking
the sum of
$\mathbf{F}_b$ over all the boundary links defining the particle.
The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
which again is summed over all links to give the total torque on the colloid.

Note that $\rho$ in the above equations is the density of the fluid
at the appropriate fluid node for the link. As the density fluctuations
in the fluid are small compared with the mean density $\rho_0$, the
density in the correction to the bounce-back can be replaced by the
mean $\rho_0$.

\subsubsection{Dynamics}
Having computed the total force and torque on an individual colloid,
is is possible to update the the linear velocities
\begin{equation}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
\end{equation}
where the mass of the colloid is related to the input radius by
$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
velocity
\begin{equation}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\mathbf{T}(t),
\end{equation}
where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
However, this explicit update is generally found to have poor stability
properties \cite{l94b, nl02}. The alternative is to use a velocity update
which is implicit \cite{heemels,nl02}.

The total force and torque on a colloid can be split into velocity-dependent
and -independent parts by combining Equations~(\ref{eq:ub})
and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
In this way, the decomposition is
\begin{eqnarray}
\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
-\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
\label{eq:d3}
\\
\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
-\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
\label{eq:d4}
\end{eqnarray}
The velocity independent parts of the force and the torque
(appropriate for a colloid at rest) are
\begin{eqnarray}
\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
\\
\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
(\mathbf{r}_b\times \mathbf{c}_b)
\end{eqnarray}
The matrices $\boldsymbol{\zeta}$ are interpreted as drag
coefficients and can be written as
\begin{eqnarray}
\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
\\
\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
(\mathbf{r}_b \times \mathbf{c}_b).
\end{eqnarray}
As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is

\begin{equation}
\label{eq:k14}
\left( \begin{array}{rrr}
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\end{array} \right).
\end{equation}
If the particle has a symmetric distribution of boundary links (the
special case where the centre of mass is on a symmetry point of the
lattice) the $\boldsymbol{\zeta}^{FU}$ and
$\boldsymbol{\zeta}^{T\Omega}$ matrices are
diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
and $\boldsymbol{\zeta}^{TU}$ matrices are zero.

The velocity updates can now be rewritten using an
implicit update, which results in
\begin{eqnarray}
\label{eq:k15}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
\\
\label{eq:k16}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
\end{eqnarray}
In general all the matrix elements are non-zero, in which case
equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
6$\times$6 matrix inversion.

Finally, the position of the particle can be updated: an Euler
forward step is taken using the updated velocity
\begin{equation}
\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
\end{equation}

\subsubsection{Changes in particle shape}

One consequence is that the discrete shape of the particle fluctuates
as the colloid centre moves relative to the
lattice. This means that the size of the particle as seen by the
fluid changes despite the fact that the input radius of the colloid
is fixed.  However, these fluctuations are small for input radii
greater than about 5 lattice units \cite{l96a}. Furthermore, the
fluctuations may be greater for some input radii than others
\cite{nl02}.

In the standard approach, changes in the map of boundary links
are accomodated by the internal fluid. If the particle shape
changes so that an internal node is exposed, the internal fluid
at that site rejoins the fluid proper with the solid body momentum
(assuming the internal fluid is relaxed to solid-body rotation).
If a fluid node is covered by particle movement, then it simply
joins the internal fluid, and will relax to solid-body rotation.
Without internal fluid, changes in particle shape
in which fluid nodes are either covered or uncovered must be
accompanied by the removal or addition
of fluid with appropriate properties at the nodes in question.
It is essential that this
is done in a way which minimises the perturbation to the fluid
flow.

If a fluid node at $\mathbf{r}$ is covered by the movement of a
solid particle,
the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
explicitly so that
the overall mass density is unchanged. At the same time, the
colloid must assume the momentum lost by the fluid
$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.

Similarly, when a fluid node is exposed by particle movement,
fluid must be replaced with the appropriate distribution, density
and velocity. If the
new density is $\rho$ (to be determined by some method) then
the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
\Delta x^3$ which must be balanced elsewhere. If the new fluid
is assumed to have velocity $\mathbf{u}$ then the particle must
give up momentum $\Delta x^3 \rho\mathbf{u}$.

>>>>

#############################################################################
#
#  porous.tex
#
#  Some considerations for porous media.
#
#  $Id: porous.tex,v 1.1 2010-11-29 16:45:57 kevin Exp $
#
#  Edinburgh Soft Matter and Statistical Physics Group and
#  Edinburgh Parallel Computing Centre
#
#  Kevin Stratford (kevin@epcc.ed.ac.uk)
#  (c) 2010 The University of Edinburgh
#
#############################################################################

@@
